Probabilistic forecasting of hourly Emergency Department arrivals

### 2022-02-10 paper/paper.Rmd
add a description for figure 8. Add title page and cover letter

5. We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements in other services, such as the number of incidents or call volumes in clinical desk services.

@@

<!--- ```{r seasonplot-dofw, fig.cap= "Seasonal plot: day of week arrivals", fig.align='center',out.width="70%"}
``` --->

Figure \@ref(fig:seasonplot-weekofyear) illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals. There is a line for each week (i.e. seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle. It is clear that there is a significant jump in patient arrivals on Monday followed by Saturday each week. This indicates that there are significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.

@@

Furthermore, we consider an extnsion to the additive models for $\lambda_t$ and $\mu_t$ above to incorporate school and public holidays into $D$. These models are labelled Poisson-2 and NOtr-2. We also performed experiments with the truncated $t$ distribution (Ttr-2) and negative binomial distribution (NBI-2), but these did not results in forecasts as well calibrated as the truncated normal.

@@

```{r Pinball, fig.cap= "Pinball loss values by quantile. Benchmark-2, NBI-2 and NOtr-2 have similar performance with most improvement of the latter two over the benchmark comming from the lower quantiles. All other models have a greater Pinball score (worse performance) than Benchmark-2 across all quantiles.", fig.align='center',out.width= "90%"}
```

@@

```{r quantile-bias, fig.cap= "Quantile bias illustrates the difference between the nominal exceedance of prediceted quantiles and the observed frequency of exceedance. If a predictive model is calibrated, quantile bias should be approximately zero accross all probability levels. Here we see that the predictive distributions produced by TBATS are underdispersed (too narrow/over confident); those produced by NBI-2, GBM-2 and Benchmark-2" bias and underpredict accross all probability levels; and that forecasts produced by ADAM-iETSX, NOtr-2 and Prophet are well calibrated.", fig.align='center',out.width= "90%"}
```

@@

Finally, Figure \@ref(fig:time-accuracy) reports the forecast accuracy of each approach presented in Table \@ref(tab:tab-results) versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours. The x-axis represents the speed of each method as slow, moderate or fast.
<!-- We need to explain this once we have th result for RMSE -->
<!-- ``` -->

@@

Two approaches produced highly accurate, calibrated probabilistic forecasts: a time series model and a model based on distributional regression. The first is ADAM-iESTX, an extension of exponential smoothing incorporating multiple seasonalities, explanatory variables and assuming a Gamma predictive distribution. The second, labelled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on the date and time features and temperature. Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while NOtr-2 produced forecasts with a lower pinball score. This suggests that the latter may be preferred if the whole distribution is used in decision-making.

>>>>

5. We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements in other services, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.

@@

 ```{r seasonplot-dofw, fig.cap= "Seasonal plot: day of week arrivals", fig.align='center',out.width="70%"}
 ```

Figure \@ref(fig:seasonplot-dofw) illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals. There is a line for each week (i.e. seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle. It is clear that there is a significant jump in patient arrivals on Monday followed by Saturday each week. This indicates that there are significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.

@@

Furthermore, we consider an extension to the additive models for $\lambda_t$ and $\mu_t$ above to incorporate school and public holidays into $D$. These models are labelled Poisson-2 and NOtr-2. We also performed experiments with the truncated $t$ distribution (Ttr-2) and negative binomial distribution (NBI-2), but these did not results in forecasts as well calibrated as the truncated normal.

@@

```{r Pinball, fig.cap= "Pinball loss values by quantile. Benchmark-2, NBI-2 and NOtr-2 have similar performance with most improvement of the latter two over the benchmark comming from the lower quantiles. All other models have a greater Pinball score (worse performance) than Benchmark-2 across all quantiles.", fig.align='center',out.width= "70%"}
```

@@

```{r quantile-bias, fig.cap= "Quantile bias illustrates the difference between the nominal exceedance of prediceted quantiles and the observed frequency of exceedance. If a predictive model is calibrated, quantile bias should be approximately zero accross all probability levels. Here we see that the predictive distributions produced by TBATS are underdispersed (too narrow/over confident); those produced by NBI-2, GBM-2 and Benchmark-2 bias and underpredict accross all probability levels; and that forecasts produced by ADAM-iETSX, NOtr-2 and Prophet are well calibrated.", fig.align='center',out.width= "70%"}
```

@@

Finally, Figure \@ref(fig:time-accuracy) reports the forecast performance of each approach presented in Table \@ref(tab:tab-results) versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours. The X-axis shows the speed of each method presented as slow, moderate or fast. We observe that there is no clear association between speed and accuracy improvement of the models used in this study. We found that most of the methods considered in this paper fall into the slow category with a different range of performance, depending on the type of error measure. Figure \@ref(fig:time-accuracy) also highlights that while NOtr-2 might be slow, but its accuracy is consistent when assessed using different accuracy measures. The figure also show that the $t$ family GAMLSS model (Ttr-2) is the fastest model with a reasonably accurate performance across the three presented measures. Therefore, if the speed is a major constraint when generating the forecasts, one may employ this approach to generate forecasts.

@@

Two approaches produced highly accurate, calibrated probabilistic forecasts: a time series model and a model based on distributional regression. The first is ADAM-iESTX, an extension of exponential smoothing incorporating multiple seasonalities, explanatory variables and assuming a Gamma predictive distribution. The second, labeled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on the date and time features and temperature. Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while NOtr-2 produced forecasts with a lower pinball score. This suggests that the latter may be preferred if the whole distribution is used in decision-making.

### 2022-02-10 paper/paper.tex
add a description for figure 8. Add title page and cover letter

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We produce probabilistic forecasts, in addition to the point estimation, quantifying uncertainties in future hospital admission, and comparing different forecasting methods using a suite of well-established evaluation metrics;
\item
  We develop an advanced dynamic model to forecast ED arrivals based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies, which produced highly-accurate point forecasts;
\item
  We develop a novel model to produce a probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and iii) variation in forecast uncertainty;
\item
  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and the simple empirical distribution of the arrivals;
\item
  We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements in other services, such as the number of incidents or call volumes in clinical desk services.
\end{enumerate}

@@

\caption{Distrubution of admission per hour and day of the week}\label{fig:hourly-plot-ridge}
\end{figure}

@@

Figure \ref{fig:seasonplot-weekofyear} illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals. There is a line for each week (i.e.~seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle. It is clear that there is a significant jump in patient arrivals on Monday followed by Saturday each week. This indicates that there are significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.

@@

\caption{Week of year arrivals}\label{fig:seasonplot-weekofyear}
\end{figure}

@@

\caption{Daily arrivals}\label{fig:date-plot}
\end{figure}

@@

Furthermore, we consider an extnsion to the additive models for \(\lambda_t\) and \(\mu_t\) above to incorporate school and public holidays into \(D\). These models are labelled Poisson-2 and NOtr-2. We also performed experiments with the truncated \(t\) distribution (Ttr-2) and negative binomial distribution (NBI-2), but these did not results in forecasts as well calibrated as the truncated normal.

@@

{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/Pinball-1}

@@

\caption{Pinball loss values over different quantiles.}\label{fig:Pinball}
\end{figure}

@@

{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/quantile-bias-1}

@@

\caption{Quantile bias vs the nominal quantiles.}\label{fig:quantile-bias}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/Skill-rel2bench-1}

\caption{Pinball skill score for the seven best performing approaches.}\label{fig:Skill-rel2bench}
\end{figure}

Figure \ref{fig:Skill-rel2bench} illustrates the trade-off between calibration and pinball skill.

@@

\caption{RMSE values over the forecast horizon.}\label{fig:lead-time-rmse}
\end{figure}

@@

\caption{Speed (Running time) vs. forecast accuracy}\label{fig:time-accuracy}
\end{figure}

Finally, Figure \ref{fig:time-accuracy} reports the forecast accuracy of each approach presented in Table \ref{tab:tab-results} versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours. The x-axis represents the speed of each method as slow, moderate or fast.

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We produce probabilistic forecasts, in addition to the point estimation, quantifying uncertainties in future hospital admission, and comparing different forecasting methods using a suite of well-established evaluation metrics;
\item
  We develop an advanced dynamic model to forecast ED arrivals based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies, which produced highly-accurate point forecasts;
\item
  We develop a novel model to produce a probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and iii) variation in forecast uncertainty;
\item
  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and the simple empirical distribution of the arrivals;
\item
  We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements in other services, such as inpatient and outpatient care services, the number of attended incidents in ambulance services, or call volumes in clinical desk services.
\end{enumerate}

@@

\caption{Distrubution of admission by hour-of-day and day-of-week. Most days have a distinct pattern, such as relatively high arrivals on Monday mornings and in the early hours of Saturday and Sunday morning.}\label{fig:hourly-plot-ridge}
\end{figure}

@@

Figure \ref{fig:seasonplot-dofw} illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals. There is a line for each week (i.e.~seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle. It is clear that there is a significant jump in patient arrivals on Monday followed by Saturday each week. This indicates that there are significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.

@@

\caption{Arrivals by week-of-year. There is an annual trend of reduced arrivals durnng the summer and winter holiday period.}\label{fig:seasonplot-weekofyear}
\end{figure}

@@

\caption{Daily arrivals over the entire dataset with moving average trend (blue line) and dates of outliers labled. Low outliers invariables correspond to public holidays.}\label{fig:date-plot}
\end{figure}

@@

Furthermore, we consider an extension to the additive models for \(\lambda_t\) and \(\mu_t\) above to incorporate school and public holidays into \(D\). These models are labelled Poisson-2 and NOtr-2. We also performed experiments with the truncated \(t\) distribution (Ttr-2) and negative binomial distribution (NBI-2), but these did not results in forecasts as well calibrated as the truncated normal.

@@

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/Pinball-1}

@@

\caption{Pinball loss values by quantile. Benchmark-2, NBI-2 and NOtr-2 have similar performance with most improvement of the latter two over the benchmark comming from the lower quantiles. All other models have a greater Pinball score (worse performance) than Benchmark-2 across all quantiles.}\label{fig:Pinball}
\end{figure}

@@

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/quantile-bias-1}

@@

\caption{Quantile bias illustrates the difference between the nominal exceedance of prediceted quantiles and the observed frequency of exceedance. If a predictive model is calibrated, quantile bias should be approximately zero accross all probability levels. Here we see that the predictive distributions produced by TBATS are underdispersed (too narrow/over confident); those produced by NBI-2, GBM-2 and Benchmark-2 bias and underpredict accross all probability levels; and that forecasts produced by ADAM-iETSX, NOtr-2 and Prophet are well calibrated.}\label{fig:quantile-bias}
\end{figure}

@@

\caption{RMSE values over the forecast horizon for forecasts issues at midnight. Attendance during the morning is more challenging to predice resulting in greater RMSE during these hours for all models.}\label{fig:lead-time-rmse}
\end{figure}

@@

\caption{Running time vs. forecast performance}\label{fig:time-accuracy}
\end{figure}

Finally, Figure \ref{fig:time-accuracy} reports the forecast performance of each approach presented in Table \ref{tab:tab-results} versus the computational time required to generate the forecast for a given forecasting horizon of 48 hours. The X-axis shows the speed of each method presented as slow, moderate or fast. We observe that there is no clear association between speed and accuracy improvement of the models used in this study. We found that most of the methods considered in this paper fall into the slow category with a different range of performance, depending on the type of error measure. Figure \ref{fig:time-accuracy} also highlights that while NOtr-2 might be slow, but its accuracy is consistent when assessed using different accuracy measures. The figure also show that the \(t\) family GAMLSS model (Ttr-2) is the fastest model with a reasonably accurate performance across the three presented measures. Therefore, if the speed is a major constraint when generating the forecasts, one may employ this approach to generate forecasts.
